Figures

### 2016-08-29 case_study_ohi/ohi_single_datasets.Rmd

  geom_hline(yintercept = 0, color = 'gray50', size = .5) +
  geom_violin(fill = 'grey80', alpha = .5,
              scale = 'width',
              draw_quantiles = c(.25, .5, .75)) +
  geom_point(position = position_jitter(w = .3), 
             aes(color = mean_n_spp), alpha = 1, size = 0.25) +
  geom_violin(fill = NA, 
              draw_quantiles = c(.25, .5, .75),
              scale = 'width') +
  labs(x = 'Scenario, compared to OHI 2016',
       y = 'Change in SPP status') +
  scale_x_discrete(labels = c('1. AquaMaps only', 
                              '2. IUCN only'))

@@

ggsave(file.path(dir_fig, 'fig3a_violinplot_ohi_maps.png'),
       height = 4.5, width = 8.7, units = 'cm', dpi = 600)
# ggsave(file.path(dir_fig, 'fig3a_violinplot_ohi_maps.png'), device = 'tiff',
#        height = 4.5, width = 8.7, units = 'cm', dpi = 600)

>>>>

  geom_hline(yintercept = 0, color = 'gray50', size = .5) +
  geom_violin(fill = 'grey80', alpha = .5,
              scale = 'width',
              draw_quantiles = c(.25, .5, .75)) +
  geom_point(position = position_jitter(w = .3), 
             color = 'black', alpha = 1, size = 0.25) +
  geom_violin(fill = NA, 
              draw_quantiles = c(.25, .5, .75),
              scale = 'width') +
  labs(x = 'Scenario, compared to OHI 2016',
       y = 'Change in SPP status') +
  scale_x_discrete(labels = c('1. AquaMaps only', 
                              '2. IUCN only'))

@@

ggsave(file.path(dir_fig, 'fig4a_violinplot_ohi_maps.png'),
       height = 4.5, width = 8.7, units = 'cm', dpi = 600)
# ggsave(file.path(dir_fig, 'fig4a_violinplot_ohi_maps.png'), device = 'tiff',
#        height = 4.5, width = 8.7, units = 'cm', dpi = 600)

### 2016-08-29 manuscript-main.Rmd
sent to BH, JA, CS

---
title: "Aligning marine species range data to better serve science and conservation"
author: Casey O'Hara^1^, Jamie C. Afflerbach^1^, Courtney Scarborough^1^, Benjamin
  S. Halpern^1,2,3^
csl: refs/journal-of-applied-ecology.csl
output:
  word_document: default
  pdf_document:
    number_sections: no
    toc: no
  html_document:
    highlight: haddock
    number_sections: no
    theme: spacelab
    toc: no
bibliography: refs/zotero_lib.bib
---

@@

1. National Center for Ecological Analysis and Synthesis, 735 State Street Suite 300, Santa Barbara CA 93101
2. Bren School of Environmental Science and Management, University of California, Santa Barbara CA 93106
3. Department of Life Sciences, Imperial College London, Silwood Park Campus, Buckhurst Rd, Ascot, West Berkshire SL5 7PY, United Kingdom

@@

_Classification:_ BIOLOGICAL SCIENCES: Sustainability Science

@@

``` {r abstract, child = 'ms_0_abst_j_appl_ecol.Rmd'}
```

@@

``` {r results_discussion, child = 'ms_3_res_disc.Rmd'}
```

>>>>

---
title: "Aligning marine species range data to better serve science and conservation"
author: Casey O'Hara^1^, Jamie C. Afflerbach^1^, Courtney Scarborough^1^, Benjamin
  S. Halpern^1,2,3^
csl: refs/plos-one.csl
output:
  word_document: default
  pdf_document:
    number_sections: no
    toc: no
  html_document:
    highlight: haddock
    number_sections: no
    theme: spacelab
    toc: no
bibliography: refs/zotero_lib.bib
---

@@

1. National Center for Ecological Analysis and Synthesis, University of California, 735 State Street Suite 300, Santa Barbara CA 93101
2. Bren School of Environmental Science and Management, University of California, Santa Barbara CA 93106
3. Department of Life Sciences, Imperial College London, Silwood Park Campus, Buckhurst Rd, Ascot, West Berkshire SL5 7PY, United Kingdom

@@

@@

``` {r abstract, child = 'ms_0_abst.Rmd'}
```

@@

``` {r results_discussion, child = 'ms_3_res_disc_concl.Rmd'}
```

### 2016-08-29 ms_0_abst_j_appl_ecol.Rmd

## Abstract

@@

Species distribution data provide the foundation for a wide range of ecological research studies and conservation management decisions, yet most species ranges remain unknown, and existing range maps often suffer from data limitations and inconsistencies.  AquaMaps and the International Union for Conservation of Nature (IUCN) are two distinct efforts to map marine species distributions at a global scale.  Together these databases represent `r format(n_spp_all, big.mark = ',')` species (`r format(p_spp_am, big.mark = ',')`% within AquaMaps, `r format(p_spp_iucn, big.mark = ',')`% within IUCN), with only `r format(n_spp_both, big.mark = ',')` shared species.  Here we examine differences in predicted species ranges between the two datasets and find that these misalignments mainly result from divergent methodologies that introduce differing frequencies of commission and omission errors.  We illustrate the scientific and management implications of these differences by repeating two recent applications - an assessment of global biodiversity within the Ocean Health Index and a global analysis of gaps in coverage of marine protected areas - and find significantly different results depending on how the two datasets are used.  Until a single, highly accurate dataset of global marine species ranges becomes available, understanding the implications of dataset differences for conservation planning and decision-making remains essential.

>>>>

## Summary

@@

1. Species distribution data provide the foundation for a wide range of ecological research studies and conservation management decisions, yet most species ranges remain unknown, and existing range maps often suffer from data limitations and inconsistencies.  AquaMaps and the International Union for Conservation of Nature (IUCN) are two distinct efforts to map marine species distributions at a global scale.  Together these databases represent `r format(n_spp_all, big.mark = ',')` species (`r format(p_spp_am, big.mark = ',')`% within AquaMaps, `r format(p_spp_iucn, big.mark = ',')`% within IUCN), with only `r format(n_spp_both, big.mark = ',')` shared species.  
2. Here we examine differences in predicted species ranges between the two datasets by comparing alignment of distribution (predicted species range common to both datasets) and area (extent of range predicted by each dataset) to identify potential sources of misalignment.
3. We find that these misalignments mainly result from divergent methodologies that introduce differing frequencies of commission and omission errors. For example, species ranges predicted by IUCN range maps frequently overextend into unsuitably deep territory, as in the case of many coral and reef-associated species; similarly, the AquaMaps species distribution model often predicts species presence far afield from known occurrences when extrapolating based on limited observations.
4. __Policy implications:__ We illustrate the scientific and management implications of these differences by repeating two recent applications - an assessment of global biodiversity within the Ocean Health Index and a global analysis of gaps in coverage of marine protected areas - and find significantly different results depending on how the two datasets are used.  Until a single, highly accurate dataset of global marine species ranges becomes available, understanding the implications of dataset differences for conservation planning and decision-making remains essential.

### 2016-08-29 ms_1_intro.Rmd

One major outcome of this body of science is the various compiled databases of species distribution maps. The two most comprehensive, widely-used global-scale repositories that predict marine species ranges throughout the world's oceans are AquaMaps [@kaschner_aquamaps:_2015] and range data from the International Union for Conservation of Nature (IUCN) [@iucn_iucn_2015].  Neither dataset claims to represent the "truth" of a species' spatial distribution, but rather each offers an understanding of the "truth", with IUCN ranges relying primarily on expert opinion and AquaMaps on model predictions.  Uncertainties inherent in method and intent inevitably drive differences in the range predictions made by each dataset: geographic range data such as IUCN range maps frequently introduce commission errors (inaccurate indication of presence), while species distribution models such as AquaMaps will likely introduce fewer commission errors but more omission errors (inaccurate indications of absence) [@rondinini_tradeoffs_2006].  Each type of error bears different implications for conservation goals: commission errors can result in prioritizing areas not relevant to conservation goals, while omission errors may result in protected area networks that fail to include important habitat and range [@rondinini_tradeoffs_2006; @jetz_ecological_2008].

These two spatial datasets have been used in hundreds of studies and applications for a wide range of purposes, including assessing marine species status [@halpern_index_2012; @halpern_patterns_2015; @selig_assessing_2013-1], evaluating global biodiversity patterns [@coll_biodiversity_2010; @martin_manual_2014; @pimm_biodiversity_2014; @kaschner_current_2011], predicting species range shifts [@garciamolinos_climate_2015], and setting conservation priorities [@klein_shortfalls_2015].  In most of these cases, the implications of choosing one versus the other of these datasets is not evaluated or discussed, yet strong conservation and management conclusions are drawn from the results of the studies. For example, biases in taxonomic or spatial coverage within a dataset could shift management and conservation actions away from places or species that are most in need.

@@

We then reexamine two recent marine biodiversity studies - an assessment of the status of global biodiversity within the Ocean Health Index (OHI) [@halpern_index_2012; @halpern_patterns_2015] and a global analysis of gaps in protection afforded by marine protected areas (MPAs) [@klein_shortfalls_2015] - as case studies to explore the implications of prioritizing one data set over the other.  The results highlight possible consequences of different data use decisions on our understanding of marine biodiversity status and protection.

>>>>

One major outcome of this body of science is the various compiled databases of species distribution maps. The two most comprehensive, widely-used global-scale repositories that predict marine species ranges throughout the world's oceans are AquaMaps [@kaschner_aquamaps:_2015] and range data from the International Union for Conservation of Nature (IUCN) [@iucn_iucn_2015].  Neither dataset claims to represent the "truth" of a species' spatial distribution, but rather each offers an understanding of the truth, with IUCN ranges relying primarily on expert opinion and AquaMaps on model predictions.  Uncertainties inherent in method and intent inevitably drive differences in the range predictions made by each dataset: geographic range data such as IUCN range maps frequently introduce commission errors (inaccurate indication of presence), while species distribution models such as AquaMaps will likely introduce fewer commission errors at the expense of more omission errors (inaccurate indications of absence) [@rondinini_tradeoffs_2006].  Each type of error bears different implications for conservation goals: commission errors can result in prioritizing areas not relevant to conservation goals, while omission errors may result in protected area networks that fail to include important habitat and range [@rondinini_tradeoffs_2006; @jetz_ecological_2008].

These two spatial datasets have been used in hundreds of studies and applications for a wide range of purposes, including assessing marine species status [@halpern_index_2012; @halpern_patterns_2015; @selig_assessing_2013-1], evaluating global biodiversity patterns [@coll_biodiversity_2010; @martin_manual_2014; @pimm_biodiversity_2014; @kaschner_current_2011], predicting species range shifts [@garciamolinos_climate_2015], and setting conservation priorities [@klein_shortfalls_2015].  In most of these cases, the implications of choosing one versus the other of these datasets is not evaluated or discussed, yet strong conservation and management conclusions are drawn from the results of the studies. <!-- For example, biases in taxonomic or spatial coverage within a dataset could shift management and conservation actions away from places or species that are most in need.-->

@@

<!--
We then reexamine two recent marine biodiversity studies - an assessment of the status of global biodiversity within the Ocean Health Index (OHI) [@halpern_index_2012; @halpern_patterns_2015] and a global analysis of gaps in protection afforded by marine protected areas (MPAs) [@klein_shortfalls_2015] - as case studies to explore the implications of prioritizing one data set over the other.  The results highlight possible consequences of different data use decisions on our understanding of marine biodiversity status and protection.  
-->

We then reexamine a global analysis of gaps in protection afforded by marine protected areas (MPAs) [@klein_shortfalls_2015] - as a case study to explore the implications of prioritizing one data set over the other.  The results highlight possible consequences of different data use decisions on our understanding of marine biodiversity status and protection.

### 2016-08-29 ms_2_methods.Rmd

___OMIT: Studies using AquaMaps data frequently define "presence" by applying a probability threshold, e.g., previously, the Ocean Health Index defined presence as 40% or greater probability of occurrence [@halpern_index_2012; @halpern_patterns_2015].  In the 2016 OHI assessment, the presence threshold was eliminated, to better approximate the "extent of occurrence" generally indicated by IUCN maps, since the two datasets are used together.___

@@

>>>>

<!--
___OMIT: Studies using AquaMaps data frequently define "presence" by applying a probability threshold, e.g., previously, the Ocean Health Index defined presence as 40% or greater probability of occurrence [@halpern_index_2012; @halpern_patterns_2015].  In the 2016 OHI assessment, the presence threshold was eliminated, to better approximate the "extent of occurrence" generally indicated by IUCN maps, since the two datasets are used together.___
-->

@@

<!--
